# scrubR: A Data Cleaning and Visualization Package for STAT 108

## Why Use This Package?

**scrubR** is an R package designed to simplify data cleaning, validation, and analysis for datasets like sports and clinical trials. It’s aimed at students, researchers, or anyone who needs easy-to-use tools for tidying data, running checks, and generating basic plots.

### Features:
- **Data Cleaning:** Fix columns, handle missing data, and parse entries like relay distances or times.
- **Data Validation:** Check for unexpected values or inconsistencies in numeric and categorical data.
- **Visualization:** Create plots to explore trends or identify issues.
- **Easy Workflows:** Especially useful for projects with frequent data updates.

If you’re analyzing Olympic swimming results or tracking mouse weights in trials, **scrubR** provides a straightforward way to prepare your data for analysis.

---

## Installation

Install **scrubR** directly from GitHub:

```{r}
# install remotes package if don’t have it
install.packages("remotes")

# install scrubR package
remotes::install_github("SDulay13/scrubR")
```

---

## How to Use This Package

### Example 1: Cleaning Swimming Data

The `clean_swim_data()` function processes Olympic swimming event results. It:
- Parses relay distances (e.g., `4x100` → `400` meters).
- Converts performance times to seconds.
- Identifies and flags problematic rows.

#### Code Example:

```{r}
library(scrubR)

# load swimming data
data <- read.csv("data-raw/Olympic_Swimming_Results_1912to2020.csv")

# clean the data
swim_data_cleaned <- clean_swim_data(data, rename_columns = TRUE, verbose = TRUE)

head(swim_data_cleaned)
```

---

### Example 2: Monitoring Weight Changes

The `plot_change()` function visualizes changes in data over time, flagging significant drops (e.g., weight loss >20%) in trials.

#### Code Example:

```{r}
# load mouse data
file_path <- "inst/extdata/mousedata.xlsx"
if (file.exists(file_path)) {
  mouse_data <- join_sheets(file_path, "ID", c("Birth", "Body Weight", "Outcome"))
} else {
  stop("file does not exist. check the path.")
}

# plot weight changes
plot <- plot_change(
  data = mouse_data,
  id_col = "ID",
  var_cols = c("Body Weight 1", "Body Weight 2", "Body Weight 3"),
  color = "Treatment",
  cat = "Treatment"
)
plot
```

---

## Package Structure

### Main Functions:

1. **`clean_swim_data()`**: Cleans and processes swimming event datasets.
2. **`categorical_check()`**: Validates categorical column entries.
3. **`quantitative_check()`**: Detects non-numeric or invalid numeric values.
4. **`plot_change()`**: Visualizes longitudinal trends and flags outliers.
5. **`join_sheets()`**: Combines multiple Excel sheets and finds mismatches.

### Example Datasets:
- [Olympic Swimming Data](https://www.kaggle.com/datasets/datasciencedonut/olympic-swimming-1912-to-2020)
- Clinical trial data for mice (e.g., tracking weight changes).

For more detailed examples, refer to the Vignette.

---

## Contributions

**Developers**: Hana Rostami, Loretta Eng, Sukhraj Dulay

We welcome feedback and contributions! Submit issues or pull requests via our [GitHub repository](https://github.com/SDulay13/scrubR).

---

```{r}
devtools::build_readme()
```
